---
title: "Forecasting Australian Real Wages"
author: "Stephan Berke"

execute:
  echo: false
  
bibliography: references.bib
---

> **Abstract.** This project employs a Bayesian Vector Autoregression (BVAR) model to forecast Australian real wages for future periods. Beginning in 2021, Australian households experienced negative growth rates in real wages, but recent trends indicate a rebound. Therefore, a positive trajectory in real wages is anticipated for the upcoming periods.
>
> **Keywords.** BVARs, Real Wages, Inflation, Households

# Research Question

How will real wages and the broader economic conditions for households in Australia evolve in the upcoming periods?

# Motivation

Accurate forecasting of real wages is imperative for policymakers and businesses as it facilitates the anticipation of shifts in consumer purchasing power, which in turn substantially influences economic demand and inflationary dynamics. Such forecasts are integral to the formulation of monetary and fiscal policies aimed at maintaining economic stability and fostering long-term growth. The rising inflation and consequent escalation in the cost of living have imposed significant strain on Australian households (@RBA2024), compounded by the inability of nominal wage growth to sustain its historical average of approximately 4% since 2013 (@ABS2024WPI). This stagnation has resulted in a marked decline in real wages, particularly in the aftermath of the COVID-19 pandemic (@TreasuryAustralia2024). Notably, there has been a discernible upsurge in real wages since the last quarter. Nevertheless, historical projections by the Reserve Bank of Australia (RBA) have frequently failed to align with actual developments (@RBA2017), highlighting the necessity for a more refined predictive algorithm that can reliably anticipate future stable increases in real wages.

## Theory base

**Real Wages** defined as:

```{=tex}
\begin{gather}
\log Real\:wages = \log Nominal\:wages\:- \log CPI 
\end{gather}
```

# Data

Labor data is retrieved from the Australian Bureau of Statistics (ABS) using the **readabs** function and the Reserve Bank of Australia (RBA) using **readrba**.

The ABS releases data on average weekly earnings biannually, in May and December As the latest data is from 4 months ago, we will utilize the Wage Price Index (WPI), which is measured quarterly (last released in December 2023).

Key economic variables are incorporated to understand labor market and economic dynamics. Nominal and Real Wages assess purchasing power and income trends, with real wages adjusted using the Consumer Price Index (CPI) to account for inflation. Producer Prices, reflected by the Producer Price Index (PPI), provide insights into production costs and business environment.Export (EPI) and Import Indexes (IPI) are included to evaluate trade impacts and economic competitiveness. The Cost of Living Index (CLI) assesses consumer expenses, influencing economic welfare. Unemployment and Labor Participation Rates offer perspectives on labor market health and engagement. Log Hours Worked and Log Real GDP are used to analyze productivity trends and overall economic output.

## Indicators

#### Australian Labor and Financial Data

Following indicators will be used in the model. Indicators on global economic factors will be included later in this project.

| Indicator           | Index     | Source  | Unit        | Period    |
|---------------------|-----------|---------|-------------|-----------|
| Real Wages          | WPI - CPI | ABS/RBA | \%          | 1997-2024 |
| Inflation           | CPI       | ABS     | \%          | 1948-2023 |
| Producer Prices     | PPI       | ABS     | \%          | 1998-2023 |
| Export Index        | EPI       | ABS     | \%          | 1998-2023 |
| Import Index        | IPI       | ABS     | \%          | 1998-2023 |
| Cost of Living      | CLI       | ABS     | \%          | 1998-2023 |
| Unemployment        | UR        | RBA     | \%          | 1978-2024 |
| Labor Participation | LPR       | ABS     | \%          | 1978-2024 |
| Log Hours Worked    | HW        | RBA     | \-          | 1978-2024 |
| Log Real GDP        | Real GDP  | RBA     | Million AUD | 1959-2023 |

## Data extraction and transformation

Since the main indicators are based on quarterly data, all indicators are converted into quarterly time series.

Given that the extracted time series data for the production price index starts in 1998 Q4, this quarter will work as the starting date of the analysis.

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(quantmod)
library(ggplot2)
library(readrba)
library(readabs)
library(dplyr)
library(xts)
library(tframePlus)
library(zoo)
library(knitr)
library(kableExtra)
library(forecast)
library(tseries)
```

```{r data extraction, message=FALSE, warning=FALSE}

#define periods
start.period <- "1998-12-01"
start.period.qt <- "1998 Q4"
end.period <- "2023-12-31"
end.period.qt <- "2023 Q4"


# Get wage growth
log.nom.wpi <- log(subset(read_abs(series_id = "A2713851R"), date >= as.Date(start.period) & date <= as.Date(end.period))$value)
                          
# Get inflation rate (CPI)
log.cpi <- log(subset(read_abs(series_id = "A2325846C"), date >= as.Date(start.period) & date <= as.Date(end.period))$value)  
                          
# Define real wage
log.real.wpi <- log.nom.wpi - log.cpi

# Get dates
dates <- subset(read_rba_seriesid("GWPIPRIQP"), date >= as.Date(start.period) & date <= as.Date(end.period))$date

# Get producer prices (PPI)
log.ppi <- log(subset(read_abs(series_id = "A2314865F"), date >= as.Date(start.period) & date <= as.Date(end.period))$value)

# Get export price index (EXPI)
log.expi <- log(subset(read_abs(series_id = "A2294886K"), date >= as.Date(start.period) & date <= as.Date(end.period))$value)

# Get import price index (IMPI)
log.impi <- log(subset(read_abs(series_id = "A2295765J"), date >= as.Date(start.period) & date <= as.Date(end.period))$value)

# Get cost of living index (CLI)
log.cli <- log(subset(read_abs(series_id = "A4083524T"), date >= as.Date(start.period) & date <= as.Date(end.period))$value)

# Get Log of Unemployment rate
log.unempl <- read_abs(series_id = "A84423092X") |> 
  select(date, value) |> 
  mutate(q_dates = as.yearqtr(date)) |> 
  filter(q_dates >= start.period.qt & q_dates <= end.period.qt) |> 
  group_by(q_dates) |> 
  summarise(value = mean(value, na.rm = TRUE)) |> 
  mutate(log_value = log(value)) |> 
  select(log_value)  |> 
  rename(log.unempl = log_value)

#Log of labor participation
log.parti <- read_abs(series_id = "A84423093A") |>
  select(date, value) |>
  mutate(q_dates = as.yearqtr(date),
    log_value = log(value)) |>
  filter(q_dates >= start.period.qt & q_dates <= end.period.qt) |>
  group_by(q_dates) |>
  summarise(mean_log_value = mean(log_value, na.rm = TRUE)) |>
  select(mean_log_value) |>
  rename(log.parti = mean_log_value)


#hours worked
log.h.worked <- read_rba_seriesid("GLFMHW") |>
  select(date, value) |>
  mutate(q_dates = as.yearqtr(date)) |>
  filter(q_dates >= start.period.qt & q_dates <= end.period.qt) |>
  group_by(q_dates) |>
  summarise(quarterly_sum = sum(value, na.rm = TRUE)) |>
  mutate(log_quarterly_sum = log(quarterly_sum)) |>
  select(log_quarterly_sum) |>
  rename(log.h.worked = log_quarterly_sum)

#real gdp
log.real.gdp <- log(subset(read_rba_seriesid("GGDPCVGDP"), date >= as.Date(start.period) & date <= as.Date(end.period))$value)

```

## Descriptive Analysis

### Data Overview

```{r HEAD, echo=FALSE}

# Create data frame
df <- data.frame(
  "Log of Real WPI" = log.real.wpi,
  "Log of CPI" = log.cpi,
  "Log of PPI" = log.ppi,
  "Log of EXPI" = log.expi,
  "Log of IMPI" = log.impi,
  "Log of CLI" = log.cli,
  "Log of Unemployment" = log.unempl,
  "Log of Participation" = log.parti,
  "Log Hours Worked" = log.h.worked,
  "Log of Real GDP" = log.real.gdp
)
# Display the first 5 rows of the dataframe

df_display <- data.frame(
  "Date" = as.Date(dates),
  df
)
kable(head(df_display, 5))
```

### Data Plots

The accompanying graphs depict the recent trends in the selected economic indicators. Notably, the Real Wage Price Index has dropped significantly since the beginning of COVID in March 2020. Concurrently, the Consumer Price Index has exhibited larger growth rates in the same period. Recently, a slight upward trend in Real Wages can be observed, though they remain below pre-pandemic levels.

Other indicators like Cost of Living Index, the Import Index and the Producer Price Index also reveal increasing trends, highlighting notable shifts in economic conditions, with Export Prices also shifting upwards. However, labor market conditions appear resilient, with falling unemployment, stable participation rates, and increased working hours. Real GDP's upward trend suggests a recovery to pre-pandemic economic activity levels.

```{r PLOTS, echo=FALSE}
# Plot data
var_names <- colnames(df)
par(mfrow = c(4, 3), mar = c(2, 2, 2, 2))

for (j in 1:ncol(df)) {
  plot(x = dates, y = df[, j], type = 'l',
       main = var_names[j], ylab = "", xlab = "",
       lwd = 2.5,
       ylim = c(min(df[, j], na.rm = TRUE), max(df[, j], na.rm = TRUE)))
}
```

## Preliminary data analysis

### ACF Analysis

The Autocorrelation Function (ACF) plots demonstrate a decay in correlation coefficients for all time series, with the memories of Import Index and Unemployment dropping slightly faster than the rest, indicating the presence of substantial memory. The initial lags exhibit significant autocorrelation, signifying non-stationary behavior. This persistence in the time series necessitates statistical differencing to ensure stationarity.

```{r ACF plots, echo = FALSE}
par(mfrow = c(4, 3), mar = c(2, 2, 3, 2))
for (j in 1:ncol(df)) {
  acf(df[,j], main=colnames(df[j]),  )
}
```

### PACF Analysis

The Partial Autocorrelation (PACF) plots for the time series predominantly exhibit significant partial autocorrelations at the first lag for all variables, followed by a rapid convergence to the confidence bounds. 
This pattern suggests that the data could be wel represented by a first order autoregrssive model. 
```{r PACF plots, echo = FALSE}
par(mfrow = c(4, 3), mar = c(2, 2, 3, 2))
for (j in 1:ncol(df)) {
  pacf(df[,j], main=colnames(df)[j])
}
```

### Augmented Dickey-Fuller test

The ADF test results confirm the non-stationarity observed in the ACF plots for most variables. Specifically, first differences suffices for all variables except the unemployment rate, which is stationary on the 5% level, and the CPI and CLI, which require second-order differencing. 
```{r ADF_table,  message=FALSE, warning=FALSE}

#code  idea from Yobin(2023)
perform_adf_tests <- function(df) {
  # Create an empty dataframe to store the results
  results <- data.frame(Variable = character(), TestType = character(), 
                        TestStatistic = numeric(), PValue = numeric(), 
                        stringsAsFactors = FALSE)

  # Iterate over each column in the dataframe
  for (col in colnames(df)) {
    # Remove NA values from the column
    column_data <- na.omit(df[[col]])

    # Perform ADF test for levels
    adf_levels <- tseries::adf.test(column_data, k = 4)

    # Check if p-value is less than or equal to 0.05
    if (adf_levels$p.value <= 0.05) {
      results <- bind_rows(results,
        data.frame(Variable = col, TestType = "Levels", 
                   TestStatistic = adf_levels$statistic,
                   PValue = adf_levels$p.value)
      )
    } else {
      # Perform ADF test for first difference
      adf_diff1 <- tseries::adf.test(diff(column_data), k = 4)
      
      # Check if p-value is less than 0.05
      if (adf_diff1$p.value < 0.05) {
        results <- bind_rows(results,
          data.frame(Variable = col, TestType = "First Difference", 
                     TestStatistic = adf_diff1$statistic,
                     PValue = adf_diff1$p.value)
        )
      } else {
        # Perform ADF test for second difference
        adf_diff2 <- tseries::adf.test(diff(column_data, differences = 2), k = 4)
        
        results <- bind_rows(results,
          data.frame(Variable = col, TestType = "Second Difference", 
                     TestStatistic = adf_diff2$statistic,
                     PValue = adf_diff2$p.value)
        )
      }
    }
  }

  # Return the results dataframe
  return(results)
}

adf_test_results <- perform_adf_tests(df)
rmarkdown::paged_table(adf_test_results, options = list(pageLength = 11))
```

# The model

To analyze the macroeconomic and financial variables and run forecasts, a vector autoregression (VAR) model can be applied, which allows for the multivariate framework of several time series. A general VAR model with $\rho$ lags for $t=1,...,T$ can be stated as:

```{=tex}
\begin{gather}
y_{t}   &=\mu_{0}+A_{1}y_{t-1}+...+A_{\rho}y_{t-\rho}+\epsilon_{t}\\
\epsilon_{t}|Y_{T-1}    &\sim iid\left(0_{N},\Sigma\right)
\end{gather}
```
Where $y_{t}=N\times1$ is a vector of observations at time $t$, $\mu_{0}=N\times1$ is a vector of constant terms, $A_{i}=N\times N$ is a vector of the autoregressive slope parameters, $\epsilon_{t}=N\times1$ is a vector of error terms, $Y_{t-1}$ is the information set and $\Sigma=N\times N$ is the covariance matrix of the error term.

# References {.unnumbered}
