---
title: "Forecasting Australian Real Wages Growth Rates"
author: "Stephan Berke"

execute:
  echo: false
  
bibliography: references.bib
---

> **Abstract.** This project employs a Bayesian Vector Autoregression (BVAR) model to forecast Australian real wages growth rates for future periods. Beginning in 2021, Australian households experienced negative growth rates in real wages, but recent trends indicate a rebound. Therefore, a positive trajectory in real wages growth rate is anticipated for the upcoming periods.
>
> **Keywords.** BVARs, Real Wages Growth Rates, Inflation, Households

# Research Question

This project aims to forecast real wages in Australia to provide an outlook for Australian households on the development in the upcoming periods.

# Motivation

The rising inflation and consequent escalation in the cost of living have imposed significant strain on Australian households (@RBA2024). Moreover, nominal wage growth rates in Australia have failed to sustain their previous levels of approximately 4% since 2013 (@ABS2024WPI). This decline in nominal wage growth has translated into a reduction in real wages for Australian households, particularly in the recent post-COVID era (@TreasuryAustralia2024). Notably, there has been a discernible upsurge in real wages since the last quarter. This project seeks to investigate whether households can anticipate further increases in real wages in the future and whether these increases are stable.

## Thoery base

**Real wages** defined as:

```{=tex}
\begin{align}
Real\:wages = \frac{Nominal\:wages}{CPI(as\:Index)} = \frac{Nominal\:wages}{1\:+CPI(as\:Rate)} =\frac{Nominal\:wages}{1\: + Inflation}
\end{align}
```

**Real Wages Rates** defined as:
```{=tex}
\begin{align}
Real\:wages = Nominal\:wages\:-\: Inflation}
\end{align}
```

## Model base

**BVARs Model** in this research:
```{=tex}
\begin{aligned}
y_t &= \mu_0 + A_1y_{t-1} + A_2y_{t-2}...+A_py_{t-p} +\epsilon_t\\

\epsilon_t|Y_{t-1} &\sim iid \mathcal{N}_{17}(0_{17}, \Sigma)
\end{aligned}
```

### Explain of model assumptions and labels

# Data

Labor data is retrieved from the Australian Bureau of Statistics (ABS) using the readabs function. Financial data is obtained from the Reserve Bank of Australia (RBA) using the readrba function.

The ABS releases data on average weekly earnings biannually, in May and November. As the latest data is from 5 months ago, we will utilize the Wage Price Index (WPI), which is measured quarterly (last released in December 2023).

## Indicators

#### Australian Labor and Financial Data

Following indicators will be used in the model. 
Indicators on global economic factors will be included later in this project.

| Indicator           | Index     | Source  | Unit        | Period    |
|---------------------|-----------|---------|-------------|-----------|
| Nominal Wages       | WPI       | RBA     | \%          | 1997-2024 |
| Inflation           | CPI       | ABS     | \%          | 1948-2023 |
| Real Wages          | WPI - CPI | ABS/RBA | \%          | 1997-2024 |
| Producer Prices     | PPI       | ABS     | \%          | 1998-2023 |
| Export Index        | EPI       | ABS     | \%          | 1998-2023 |
| Import Index        | IPI       | ABS     | \%          | 1998-2023 |
| Cost of Living      | CLI       | ABS     | \%          | 1998-2023 |
| Unemployment        | UR        | RBA     | Individuals | 1978-2024 |
| Labor Participation | LPR       | ABS     | \%          | 1978-2024 |
| Log Real GDP        | Real GDP  | RBA     | Million AUD | 1959-2023 |

## Data extraction and transformation


Since our main indicators are based on quarterly data, all indicators are converted into quarterly time series.

Given that the extracted time series data for the production price index starts in 1998 Q4, we will use this as the starting date for our analysis.

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(quantmod)
library(ggplot2)
library(readrba)
library(readabs)
library(dplyr)
library(xts)
library(tframePlus)
library(zoo)
library(knitr)
library(kableExtra)
library(forecast)
library(tseries)
```


```{r data extraction, message=FALSE, warning=FALSE}

#define periods
end.period <- "2023-12-31"
start.period <- "1998-12-01"

# Get wage growth
nom.wpi <- subset(read_rba_seriesid("GWPIPRIQP"), date >= as.Date(start.period))$value

# Get inflation rate (CPI)
cpi <- subset(read_abs(series_id = "A2325850V"), date >= as.Date(start.period))$value

# Get dates
dates <- subset(read_rba_seriesid("GWPIPRIQP"), date >= as.Date(start.period))$date

# Define real wage
real.wpi <- nom.wpi - cpi

# Get producer prices (PPI)
ppi <- subset(read_abs(series_id = "A2314866J"), date >= as.Date(start.period))$value

# Get export price index (EXPI)
expi <- subset(read_abs(series_id = "A2294887L"), date >= as.Date(start.period))$value

# Get import price index (IMPI)
impi <- subset(read_abs(series_id = "A2295766K"), date >= as.Date(start.period))$value

# Get cost of living index (CLI)
cli <- subset(read_abs(series_id = "A4083524T"), date >= as.Date(start.period))$value

#unemployment data
unemployment.data <- read_rba_seriesid("GLFSURSA")
unempl.yl_xts <- xts(unemployment.data$value, order.by = as.Date(unemployment.data$date))
unempl.qt <- to.quarterly(unempl.yl_xts, OHLC = FALSE)
unempl.qt_filtered <- window(unempl.qt, start = as.yearqtr("1998 Q4"), end = as.yearqtr("2023 Q4"))
unempl <- coredata(unempl.qt_filtered)

#labor participation
participation.data <- read_abs(series_id = "A84423093A")
parti.yl_xts <- xts(participation.data$value, order.by = as.Date(participation.data$date))
parti.qt <- to.quarterly(parti.yl_xts, OHLC = FALSE)
parti.qt_filtered <- window(parti.qt, start = as.yearqtr("1998 Q4"), end = as.yearqtr("2023 Q4"))
parti <- coredata(parti.qt_filtered)


#hours worked
h.worked.data <- read_rba_seriesid("GLFMHW")
h.worked_xts <- xts(h.worked.data$value, order.by = as.Date(h.worked.data$date))
h.worked.qt <- to.quarterly(h.worked_xts, OHLC = FALSE)
h.worked.qt_filtered <- window(h.worked.qt, start = as.yearqtr("1998 Q4"), end = as.yearqtr("2023 Q4"))
h.worked <- coredata(h.worked.qt_filtered)
log.h.worked <- log(h.worked)

#real gdp
log.real.gdp <- log(subset(read_rba_seriesid("GGDPCVGDP"), date >= as.Date(start.period))$value)

```

```{r labels, echo=FALSE}

attr(nom.wpi, "label") <- "Nominal WPI"
attr(cpi, "label") <- "CPI"
attr(dates, "label") <- "Dates"
attr(real.wpi, "label") <- "Real WPI"
attr(ppi, "label") <- "PPI"
attr(expi, "label") <- "EXPI"
attr(impi, "label") <- "IMPI"
attr(cli, "label") <- "CLI"
attr(unempl, "label") <- "Unemployment"
attr(parti, "label") <- "Participation"
attr(log.h.worked, "label") <- "Log Hours Worked"
attr(log.real.gdp, "label") <- "Log Real GDP"

```


## Descriptive Analysis

### Data Overview

```{r HEAD, echo=FALSE}

# Create dataframe
df <- data.frame(
  "Nominal WPI" = nom.wpi,
  "CPI" = cpi,
  "Real WPI" = real.wpi,
  "PPI" = ppi,
  "EXPI" = expi,
  "IMPI" = impi,
  "CLI" = cli,
  "Unemployment" = unempl,
  "Participation" = parti,
  "Log Hours Worked" = log.h.worked,
  "Log Real GDP" = log.real.gdp
)
# Display the first 5 rows of the dataframe

df_display <- data.frame(
  "Date" = as.Date(dates),
  df
)
kable(head(df_display, 5))
```

### Data Plots

The following plots visualize the recent development of the variables of interest. The WPI started to climb recently, with the CPI decreasing at the same time. The real WPI finally begins to recover from negative growth rates in the past post-COVID era. 
Of particular note are the stronger growth rates observed for the cost of living and the producer price index.

```{r PLOTS, echo=FALSE}
# Plot data
var_names <- colnames(df)
par(mfrow = c(4, 3), mar = c(2, 2, 2, 2))

for (j in 1:ncol(df)) {
  plot(x = dates, y = df[, j], type = 'l',
       main = var_names[j], ylab = "", xlab = "",
       lwd = 2.5,
       ylim = c(min(df[, j], na.rm = TRUE), max(df[, j], na.rm = TRUE)))
}
```
## Preliminary data analysis

### ACF Analysis

```{r ACF plots, echo = FALSE}
par(mfrow = c(4, 3), mar = c(2, 2, 3, 2))
for (j in 1:ncol(df)) {
  acf(df[,j], main=colnames(df[j]),  )
}
```

### PACF Analysis

```{r PACF plots, echo = FALSE}
par(mfrow = c(4, 3), mar = c(2, 2, 3, 2))
for (j in 1:ncol(df)) {
  pacf(df[,j], main=colnames(df)[j])
}
```
# References {.unnumbered}
