---
title: "Real Wage Growth Rate Forecasts"
author: "Stephan Berke"

execute:
  echo: false
  
bibliography: references.bib
---

> **Abstract.** This project uses a BVAR approach to forecast Australian real wage growth rates
>
> **Keywords.** bvars, impulse responses, quarto, R, real wage growth, inflation

# Motivation


Since 2013, nominal wage growth rates in Australia have been unable to maintain their previous levels. Concurrently, households have experienced rising inflation, which has resulted in a decline in real wage growth rates in the recent post-COVID era.


# Research Question

Rising inflation and thus rising cost of living expenses have strained Australian households. This project aims to forecast real wage growth in Australia to provide an outlook on the development of real wages.    

## Model setup

We will define real wage growth as

\center{\textbf{Real Wage Growth (%) = Nominal Wage Growth (%) - Inflatoin Rate (%)}}


```{r load packages, include=FALSE, warning=FALSE, eval=FALSE}
rm(list = ls())
pkgs <- c('ggplot2', 'readrba', 'readabs' )
#install.packages(pkgs)
lapply(pkgs, library, character.only=TRUE)
```


# Data

This model utilizes data from the Reserve Bank of Australia and the Australian Bureau of Statistics.
The data is accessed via the R commands  `readrba` and `readabs`.

# Variables

Y - Real wage growth consisting of nominal wage growth  and inflation rate

X - Variables

 * Australian Labor Data 
    + Unemployment Rate 
    + Labor productivity 
    + Labor participation rate
    + Minimum Wages rate
    + Unionization
    + Demographics
  
 * Global macroeconomic indicators
    + Global GDP growth
    + Technological advances
    + Trade balances
    
```{r data extraction, include=FALSE, warning=FALSE, eval=FALSE}

#get data 
#wage growth 
wage.growth.data <- read_rba_seriesid("GWPIQP")
nom.wage.growth <- subset(wage.growth.data, date >= as.Date("1997-12-31"))$value

#inflation rate
inflation.rate.data <- read_abs(series_id = "A2325850V")
nom.inflation.rate <- subset(inflation.rate.data, date >= as.Date("1997-12-01"))$value

#dates
dates <- wage.growth.data$date

#define real wage growth
real.wage.growth <-nom.wage.growth - nom.inflation.rate

```

```{r data extraction, include=FALSE, warning=FALSE, eval=FALSE}
#create dataset
data <- data.frame(dates = dates, real.wage.growth = real.wage.growth)
```


## Plot Real Growth Rate Over Time
```{r}
ggplot(data, aes(x = dates, y = real.wage.growth)) +
  geom_line() + 
  labs(x = "Date", y = "Real Wage Growth (%)") +
  theme_minimal()
```


## References {.unnumbered}
